### 2014-05-28 paper/paper.tex
last version
changes after proof read

\author{Diego Perez, {\it Student Member, IEEE}, Sanaz Mostaghim {Member, IEEE}, Spyridon Samothrakis, {\it Student Member, IEEE}, Simon M. Lucas, {\it Senior Member, IEEE}}

@@

\begin{figure} [!t]
	\begin{center}
	\includegraphics[scale=0.235]{img/moopDec.png}
	\caption{Decision and objective spaces in a MOOP with two variables ($x_1$ and $x_2$) and two objectives ($f_1$ and $f_2$). In the objective space, yellow dots are non-optimal objective vectors, while blue dots form a non-dominated set.}
	\label{fig:moop}
	\end{center}
\end{figure}

There are many different algorithms in the litrature which are proposed to tackle multi-objective optimization problems~\cite{Deb2001}. Weighted-sum approach is one of the traditional methods in which the objectives are weighted according to user preference. The sum of the weighted objectives builds one objective function which is to be optimized. The solution to this single-objective problem is one certain solution, ideally on a Pareto-front. By varying the weights, it is possible to converge to different optimal solutions. Such linear scalarization approaches fail to find optimal solutions for problems with non-convex-shape Pareto-fronts~\cite{Deb2001}.

A popular choice for multi-objective optimization problems are evolutionary multi-objective optimization (EMO) algorithms~\cite{Coello1999, Zhou2011}. 
The goal of EMO algorithms is to find a set of optimal solutions with a good convergence to the Pareto-front as well as a good spread and diversity along the Pareto-front.
One of the most well-known algorithms in the literature is the Non-dominated Sorting Evolutionary Algorithm II (NSGA-{II})~\cite{Deb2002}, pseudocode of which is shown in Algorithm~\ref{alg:nsga2}. As in any evolutionary algorithm, NSGA-{II} evolves a set of $N$ individuals in a population denoted as $Pop$ with the difference that they are ranked according to dominance criterion and a crowding distance measure which is used to maintain a good diversity of solutions.

@@

The three main pillars of the NSGA-2 algorithm are:
\begin{itemize}
\item Fast non-dominated sorting: The function $F = \Call{fastNonDominatedSort}{U(t)}$ performs a non-dominated sorting and ranks the individuals stored in a set $U$ into several non-dominated fronts denoted as $F_i$, where the solutions in $F_0$ are the non-dominated solutions in the entire set. $F_i$ is the set of non-dominated solutions in $R \setminus (F_0 \cup \cdots \cup F_{i-1})$. 
\item Crowding distance: The function $\Call{crowdingDistanceAssignment}{F_i}$ assigns to each one of the individuals in $F_i$ a crowding distance value, which is the distance between the individual and its neighbours. The individuals with the smallest crowding distances are selected with a lower probability than the ones with the larger values.  
\item Elitism: The solutions in the first front $F_0$ always survive to the next population. 
\end{itemize}

@@

The dynamics of an RL problem are usually captured by a Markov Decision Process (MDP) which is a tuple $(S, A, T, R)$, where $S$ is the set of possible states in the problem (or game), and $s_0$ is the initial state. $A$ is the set of available actions the agent can make at any given time, and the transition model $T(s_i, a_i, s_{i+1})$ determines the probability of reaching the state $s_{i+1}$ when action $a_i$ is applied in state $s_i$. The reward function $R(s_i)$ provides a single value (\textit{reward}) that the agent must optimize, representing the desirability of the state $s_i$ reached. Finally, a policy $\pi(s_i) = a_i$ determines which actions $a_i$ must be chosen from each state $s_i \in S$\footnote{This is a deterministic policy and only valid during acting, not learning}. One of the most important challenges in RL, as shown in Section~\ref{sec:mcts}, is the trade-off between exploration and exploitation while trying to act. While learning, a policy must choose between selecting the actions that provided good rewards in the past and exploring new parts of the search space by selecting new actions.

Multi-objective Reinforcement Learning (MORL)~\cite{Vamplew2010} changes this definition by using a vector $R = {r_1, \dots, r_m}$ as rewards of the problem (instead of a scalar). Thus, MORL problems differ from RL in having more than one objective (here $m$) that must be maximized. If the objectives are independent or non-conflicting, scalarization approaches such as weighted sum approach, could be suitable to tackle the problem. Essentially, this would mean using a conventional RL algorithm on a single objective where the global reward is obtained from a weighted-sum of the multiple rewards. However, this is not always the case, as usually the objectives are conflicting and the policy $\pi$ must balance among them.

@@

The proposed algorithm was employed successfully in two domains: the DST and the Grid Scheduling problem, matching state-of-the-art results in both cases, at the expense of a high computational cost.

As mentioned before, the objective of this paper is to propose an MO-MCTS algorithm that is suitable for real-time domains. Although the work discussed here is influenced by Wang's approach, some modifications need to be done in order to overcome the high computational cost involved in their approach.

@@

Here, if $\overline{r}$ is not dominated by the local Pareto front, it is added to the front and $\overline{r}$ is propagated to its parent. In case $\overline{r}$ is dominated by the node's Pareto front, the local Pareto front does not change and there is no need to keep this propagation up the tree. Three observations can be made about the mechanism described here:

@@

As shown before, it is straightforward to discover which actions lead to what points in the front given as a solution: first gene in the best NSGA-II individual, a root's child in MO-MCTS. Hence, by identifying the point in the Pareto front that the algorithm should converge to, 
it is possible to execute the action that leads to that point.

@@

Note that, in the vanilla MCTS, there is no Pareto front obtained as a solution. Typically, in this case, rewards are calculated as a weighted sum of the objectives and a weight vector $W$, and then the action is chosen following any of the mechanisms usually employed in the literature: the action taken more often from the root; the one that leads to the best reward found; the move with the highest expected reward; or the action that maximizes Equation~\ref{eq:ucb1} in the root.

@@

The ship can perform four different moves: \textit{up}, \textit{down}, \textit{right}  and \textit{left}. If the action applied takes the ship off the grid or into the sea floor, the vessel's position does not change.  There are two objectives in this game: the number of moves performed by the ship, that must be minimized, and the value of the treasure found, that should be maximized. As can be seen in Figure~\ref{fig:dstBase}, the most valuable treasures are at a greater distance from the initial position, so the objectives are in conflict.

@@

The DST is a problem specially suited for the use of Transposition Tables (TT)~\cite{Childs2008} within MCTS. TT is a technique used to optimize three search based algorithms when two or more states can be found at different locations in the tree. It consists of sharing information between these equivalent states in a centralized manner, in order to avoid managing these positions as completely different states. Figure~\ref{fig:dstTable} shows an example of this situation in the DST.

@@

TT tables are implemented in the MCTS algorithms tested in this benchmark by using hash tables that store a \textit{representative} node for each pair $(position, moves)$ found.  The key of the hash map needs then to be obtained from three values: position coordinates $x$ and $y$ of the ship in the board, and number of moves, indicated by the depth of the node in the tree. Hence, transpositions can only happen at the same depth within the tree, a feature that has been successfully tried before in the literature~\cite{Kozelek2009}.

@@

As DST has two different objectives, number of moves and value of the treasure, the quality of a state can be assessed by two rewards, $\rho_p$ and $\rho_v$, for each objective respectively. $\rho_p$ is adjusted to be maximized using the maximum number of moves in the game, while $\rho_v$ is just the value in the cell with the treasure. Equation~\ref{eq:dst} summarizes these rewards:

@@

This needs to be done while optimizing three different objectives:  1) \textbf{Time}: the player must collect all waypoints scattered around the maze in as few time steps as possible; 2) \textbf{Fuel}: the fuel consumption by the end of the game must be minimized; and 3) \textbf{Damage}: the ship should end the game with as little damage as possible.

@@

Regarding the third objective, the ship can suffer damage in two different ways: by colliding with obstacles and driving through lava. In the former case, the ship can collide with normal obstacles (that subtract $10$ units of damage) and specially damaging obstacles ($30$ units). In the latter, lava lakes are abundant in MO-PTSP levels and, in contrast with normal surfaces, they deal one unit of damage for each time step the ship spends over this type of surface. All these subtractions are deducted from an initial counter of $5000$ points of damage.

@@

All algorithms tested in the MO-PTSP in this research employ macro-actions, a concept that can be used for coarsening the action space by applying the action chosen by the control algorithm in several consecutive cycles, instead of just in the next one.

@@

$\lambda_t$ is the fuel consumed so far, and $\lambda_0$ is the initial fuel at the start of the game. $\alpha$ is a value that balances between the fuel component and the time objective from Equation~\ref{eq:dist}. Note that, as waypoints need to keep being visited, it is necessary to include a distance measure for this reward. Otherwise, an approach that prioritizes this objective would not minimize distance to waypoints at all (the ship could just stand still: no fuel consumed is optimal), and therefore could not complete the game. The value of $\alpha$ has been determined empirically, in order to provide a good balance between these two components, and it is set to $0.66$.

Finally, Equation~\ref{eq:dam} gives the method to calculate the damage objective, $\rho_d$:

@@

$g_t$ is the damage suffered so far in the game, and $g_M$ is the maximum damage the ship can take. In this case, three different variables are used to regulate the behaviour of this objective: $\gamma$, $\beta_{1}$ and $\beta_{2}$. Both $\beta_{1}$ and $\beta_{2}$ have the same role as $\alpha$ in Equation~\ref{eq:fuel}: they balance between the time objective and the damage measure. The difference is that $\beta_{1}$ is used in high speeds, while $\beta_{2}$ is employed with low velocities. This is distinguished by the parameter $\gamma$, that can be seen as a threshold value for the ship's speed ($sp$). This differentiation is made in order to avoid low speeds in lava lakes, as this increase the damage suffered significantly. The values for these variables have also been determined empirically, and they are set to $\gamma = 0.8$, $\beta_{1} = 0.75$ and $\beta_{2} = 0.25$.

It is important to highlight again that these three rewards, $\rho_t$, $\rho_f$ and $\rho_d$, are the same for the three algorithms tested in the experiments.

@@

The experiments performed in this research compare three different algorithms in the two benchmarks presented in Section~\ref{sec:bench}: a single objective MCTS (referred to here simply as \textit{MCTS}), the Multi-Objective MCTS (\textit{MO-MCTS}) and a rolling horizon version of the NSGA-II algorithm described in Section~\ref{sec:moo} (\textit{NSGA-II}). This NSGA-II version evolves a population where the individuals are sequences of actions (macro-actions in the MO-PTSP case), obtaining the fitness from the state of the game after applying the sequence. The populations sizes, determined empirically, were set to $20$ and $50$ individuals for DST and MO-PTSP respectively. The value of $C$ in Equation~\ref{eq:ucb1} and~\ref{eq:moucb} is set to $\sqrt{2}$.

All algorithms have a limited number of evaluations before providing an action to perform. In order for these games to be real-time, the time budget allowed is close to $40$ milliseconds. With the objective of avoiding congestion peaks at the machine where the experiments are run, the average number of evaluations possible in $40$ ms is calculated and employed in the tests. This leads to $4500$ evaluations in the DST, and $500$ evaluations for MO-PTSP, using the same server where the PTSP and MO-PTSP competitions were run\footnote{Intel Core i5, 230GHz, 4GB of RAM.}.

@@

The results show clearly how approximating the optimal Pareto front allows for finding all possible solutions. Both NSGA-II and MO-MCTS approaches are able to converge to any solution in the front given the appropriate weight vector. It is important to highlight that these two algorithms employed the euclidean distance action selection (see Section~\ref{sec:momcts}). Other experiments, not included in this paper, showed that weighted sum action selection provides similar results to MCTS (that is, convergence to the edges of the front). The crucial distinction to make here is that both algorithms, NSGA-II and MO-MCTS, allow for better action selection mechanisms, that are able to overcome this problem by approximating a global Pareto front.

@@

The same three algorithms have been tested in the MO-PTSP domain. The experiments are performed in the $10$ maps available within the framework~\footnote{Available at \url{www.ptsp-game.net}}. In this case, the optimal Pareto front is not known in advance, and normally it will differ from one map to another. Hence, the mechanisms to compare the performance of the algorithms tested need to be different than the one used for DST.

The idea is as follows. First of all, $4$ different weight vectors are tested: $W_1 = (0.33,0.33,0.33)$, $W_2 = (0.1,0.3,0.6)$, $W_3 = (0.1,0.6,0.3)$ and $W_4 = (0.6,0.1,0.3)$, where each $w_i$ corresponds to the weight for an objective ($w_t$ for time, $w_f$ for fuel and $w_d$ for damage, in this order). $W_1$ treats all objectives with the same weight, while the other three give more relevance to different objectives. These vectors then provide a wide spectrum for weights in this benchmark. In this case, MO-MCTS uses these weights to select an action based on a weighted sum, which in preliminary experiments has shown better performance than the Euclidean distance mechanism.

@@

Some exceptions can be explained by analysing the map: in map 1, a fuel canister is always collected near the end, restoring the fuel level to its maximum. This leaves too few cycles to make a difference in the controller (note also that this is the map with the smallest fuel consumption). Also, in map 10, there is no difference in the damage objective: but map 10 is a map with no obstacles to damage the ship (thus only lava lakes deal damage). This also results in this map being the one with the lowest overall damage. It can also be seen that, in those maps where time has priority, the results for this objective are not as dominant as the other two. This can be explained by the fact that the time heuristic is actually part of the fuel and damage heuristics (Equation~\ref{eq:dist} in Eqs.~\ref{eq:fuel} and~\ref{eq:dam}).

@@

Extending this comparison to all maps, each pair of algorithms ends with a triplet $(D, \O, d)$, where $D$ is the number of maps where the first algorithm dominates the second, $\O$ is the amount of maps where no dominance can be established, and $d$ states the number of maps where the first is dominated by the second. Table~\ref{tab:resMOPTSP} summarizes these results for all algorithms tested.

\begin{table*}[!t]
\begin{center}
\begin{tabular}{|c|c|c|c|c|c|}
\hline
  & \textbf{$W: (w_t, w_f, w_d)$} & \textbf{MO-MCTS} \bm{$(D, \O, d)$} &  \textbf{MCTS} \bm{$(D, \O, d)$} & \textbf{NSGA-II} \bm{$(D, \O, d)$} & \textbf{PurofMovio} \bm{$(D, \O, d)$} \\ 
\hline
 \multirow{3}{*}{\textbf{MO-MCTS}} & $W_1: (0.33,0.33,0.33)$ & \multirow{3}{*}{$-$} & $(8,2,0)$ & $(8,2,0)$ & $(0,5,5)$\\
 & $W_2: (0.1,0.3,0.6)$ & & $(10,0,0)$ & $(4,6,0)$ & $(0,6,4)$\\
 & $W_3: (0.1,0.6,0.3)$ & & $(8,2,0)$ & $(7,3,0)$ & $(0,5,5)$\\
 & $W_4: (0.6,0.1,0.3)$ & & $(10,0,0)$ & $(3,7,0)$ & $(0,3,7)$\\
\hline
 \multirow{3}{*}{\textbf{MCTS}} & $W_1: (0.33,0.33,0.33)$& $(0,8,2)$ & \multirow{3}{*}{$-$} & $(0,2,8)$ & $(0,2,8)$\\
 & $W_2: (0.1,0.3,0.6)$ & $(0,0,10)$ & & $(4,2,4)$ & $(0,3,7)$\\
 & $W_3: (0.1,0.6,0.3)$ & $(0,2,8)$ & & $(0,1,9)$ & $(0,6,4)$\\
 & $W_4: (0.6,0.1,0.3)$ & $(0,0,10)$ &  & $(3,3,4)$ & $(0,1,9)$\\
\hline
 \multirow{3}{*}{\textbf{NSGA-II}} & $W_1: (0.33,0.33,0.33)$& $(0,2,8)$ & $(8,2,0)$ & \multirow{3}{*}{$-$} & $(0,4,6)$\\
 & $W_2: (0.1,0.3,0.6)$ & $(0,6,4)$ & $(4,2,4)$ & & $(0,4,6)$\\
 & $W_3: (0.1,0.6,0.3)$ & $(0,3,7)$ & $(9,1,0)$ & & $(0,5,5)$\\
 & $W_4: (0.6,0.1,0.3)$ & $(0,7,3)$ & $(4,3,3)$ &  & $(0,4,6)$\\
\hline
 \multirow{3}{*}{\textbf{PurofMovio}} & $W_1: (0.33,0.33,0.33)$& $(5,5,0)$ & $(8,2,0)$ & $(6,4,0)$ & \multirow{3}{*}{$-$}\\
 & $W_2: (0.1,0.6,0.3)$ & $(4,6,0)$ & $(7,3,0)$ & $(6,4,0)$ & \\
 & $W_3: (0.1,0.3,0.6)$ & $(5,5,0)$ & $(6,4,0)$ & $(5,5,0)$ & \\
 & $W_4: (0.6,0.1,0.3)$ & $(7,3,0)$ & $(9,1,0)$ & $(6,4,0)$ & \\
\hline
\end{tabular}
\caption{Results in MO-PTSP: Each cell indicates the triplet $(D, \O, d)$, where $D$ is the number of maps where the row algorithm dominates the column one, $\O$ is the amount of maps where no dominance can be established, and $d$ states the number of maps where the row algorithm is dominated by the column one. All algorithms followed the same route (order of waypoints and fuel canisters) in every map tested.}
\label{tab:resMOPTSP}
\end{center}
\end{table*}

One of the first things to notice is that MO-MCTS dominates MCTS and NSGA-II in most of the maps, and it is never dominated in any. Particularly, the dominance of MO-MCTS over MCTS is outstanding, even dominating in all $10$ maps for two of the weight vectors. MO-MCTS also dominates NSGA-II in more maps than in those where there is no dominance, and neither is it ever dominated by NSGA-II in any map.

@@

There is another aspect that can be further improved in the MO-PTSP benchmark, and is also applicable to other domains. It is naive to think that a unique weight vector will be the ideal one for the whole game. Specifically in the MO-PTSP, there are regions of the map where there are more obstacles or lava lakes, hence the ship is most likely to suffer higher damage there. Also, the route followed during the game affects the relative ideal speed between waypoints, or perhaps a fuel canister is going to be picked up, which will affect how the fuel objective can be managed. In general, many real-time games go through different phases, with different objectives and priorities.

>>>>

\author{Diego Perez, {\it Student Member, IEEE}, Sanaz Mostaghim, {\it Member, IEEE}, Spyridon Samothrakis, {\it Student Member, IEEE}, Simon M. Lucas, {\it Senior Member, IEEE}}

@@

\begin{figure} [!t]
	\begin{center}
	\includegraphics[scale=0.235]{img/moopDec.png}
	\caption{Decision and objective spaces in a MOP with two variables ($x_1$ and $x_2$) and two objectives ($f_1$ and $f_2$). In the objective space, yellow dots are non-optimal objective vectors, while blue dots form a non-dominated set.}
	\label{fig:moop}
	\end{center}
\end{figure}

There are many different algorithms in the literature that are proposed to tackle multi-objective optimization problems~\cite{Deb2001}. A weighted-sum approach is one of the traditional methods in which the objectives are weighted according to user preference. The sum of the weighted objectives builds one objective function to be optimized. The solution to this single-objective problem is one certain solution, ideally on a Pareto-front. By varying the weights, it is possible to converge to different optimal solutions. Such linear scalarization approaches fail to find optimal solutions for problems with non-convex-shape Pareto-fronts~\cite{Deb2001}.

A popular choice for multi-objective optimization problems are evolutionary multi-objective optimization (EMO) algorithms~\cite{Coello1999, Zhou2011}. 
The goal of EMO algorithms is to find a set of optimal solutions with a good convergence to the Pareto-front as well as a good spread and diversity along the Pareto-front.
One of the most well-known algorithms in the literature is the Non-dominated Sorting Evolutionary Algorithm II (NSGA-{II})~\cite{Deb2002}, the pseudocode of which is shown in Algorithm~\ref{alg:nsga2}. As in any evolutionary algorithm, NSGA-{II} evolves a set of $N$ individuals in a population denoted as $Pop$ with the difference that they are ranked according to a dominance criterion and a crowding distance measure, which are used to maintain a good diversity of solutions.

@@

The three main pillars of the NSGA-2 algorithm are:
\begin{itemize}
\item Fast non-dominated sorting: The function $F = \Call{fastNonDominatedSort}{U(t)}$ performs a non-dominated sorting and ranks the individuals stored in a set $U$ into several non-dominated fronts denoted as $F_i$, where the solutions in $F_0$ are the non-dominated solutions in the entire set. $F_i$ is the set of non-dominated solutions in $R \setminus (F_0 \cup \cdots \cup F_{i-1})$. 
\item Crowding distance: The function $\Call{crowdingDistanceAssignment}{F_i}$ assigns to each one of the individuals in $F_i$ a crowding distance value, which is the distance between the individual and its neighbours. The individuals with the smallest crowding distances are selected with a lower probability than the ones with larger values.  
\item Elitism: The solutions in the first front $F_0$ always survive to the next population. 
\end{itemize}

@@

The dynamics of an RL problem are usually captured by a Markov Decision Process (MDP) which is a tuple $(S, A, T, R)$, in which $S$ is the set of possible states in the problem (or game), and $s_0$ is the initial state. $A$ is the set of available actions the agent can make at any given time, and the transition model $T(s_i, a_i, s_{i+1})$ determines the probability of reaching the state $s_{i+1}$ when action $a_i$ is applied in state $s_i$. The reward function $R(s_i)$ provides a single value (\textit{reward}) that the agent must optimize, representing the desirability of the state $s_i$ reached. Finally, a policy $\pi(s_i) = a_i$ determines which actions $a_i$ must be chosen from each state $s_i \in S$\footnote{This is a deterministic policy and only valid during acting, not learning}. One of the most important challenges in RL, as shown in Section~\ref{sec:mcts}, is the trade-off between exploration and exploitation while trying to act. While learning, a policy must choose between selecting the actions that provided good rewards in the past and exploring new parts of the search space by selecting new actions.

Multi-objective Reinforcement Learning (MORL)~\cite{Vamplew2010} changes this definition by using a vector $R = {r_1, \dots, r_m}$ as rewards of the problem (instead of a scalar). Thus, MORL problems differ from RL in having more than one objective (here $m$) that must be maximized. If the objectives are independent or non-conflicting, scalarization approaches such as the weighted sum approach, could be suitable to tackle the problem. Essentially, this would mean using a conventional RL algorithm on a single objective where the global reward is obtained from a weighted-sum of the multiple rewards. However, this is not always the case, as usually the objectives are conflicting and the policy $\pi$ must balance among them.

@@

The proposed algorithm was employed successfully in two domains: the DST and the Grid Scheduling problem. In both cases, the results matched the state-of-the-art, albeit at the expense of a high computational cost.

As mentioned before, the objective of this paper is to propose an MO-MCTS algorithm that is suitable for real-time domains. Although the work discussed here is influenced by Wang's approach, some modifications need to be made in order to overcome the high computational cost involved in their approach.

@@

Here, if $\overline{r}$ is not dominated by the local Pareto front, it is added to the front and $\overline{r}$ is propagated to its parent. If $\overline{r}$ is dominated by the node's Pareto front, the local Pareto front does not change and there is no need to maintain this propagation up the tree. Three observations can be made about the mechanism described here:

@@

As shown before, it is straightforward to discover which actions lead to what points in the front given as a solution: the first gene in the best NSGA-II individual, a root's child in MO-MCTS. Hence, by identifying the point in the Pareto front that the algorithm should converge to, it is possible to execute the action that leads to that point.

@@

Note that, in the vanilla MCTS, there is no Pareto front obtained as a solution. Typically, in this case, rewards are calculated as a weighted sum of the objectives and a weight vector $W$. The action is then chosen following any of the mechanisms usually employed in the literature: the action taken more often from the root; the one that leads to the best reward found; the move with the highest expected reward; or the action that maximizes Equation~\ref{eq:ucb1} in the root.

@@

The ship can perform four different moves: \textit{up}, \textit{down}, \textit{right}  and \textit{left}. If the action applied takes the ship off the grid or into the sea floor, the vessel's position will not change.  There are two objectives in this game: the number of moves performed by the ship, which must be minimized, and the value of the treasure found, which should be maximized. As can be seen in Figure~\ref{fig:dstBase}, the most valuable treasures are at a greater distance from the initial position, so the objectives are in conflict.

@@

The DST is a problem especially suited for the use of Transposition Tables (TT)~\cite{Childs2008} within MCTS. TT is a technique used to optimize tree search-based algorithms when two or more states can be found at different locations in the tree. It consists of sharing information between these equivalent states in a centralized manner, in order to avoid managing these positions as completely different states. Figure~\ref{fig:dstTable} shows an example of this situation in the DST.

@@

TT tables are implemented in the MCTS algorithms tested in this benchmark by using hash tables that store a \textit{representative} node for each pair $(position, moves)$ found.  The key of the hash map then needs to be obtained from three values: position coordinates $x$ and $y$ of the ship in the board, and number of moves, indicated by the depth of the node in the tree. Hence, transpositions can only happen at the same depth within the tree, a feature that has been successfully tried before in the literature~\cite{Kozelek2009}.

@@

As DST has two different objectives, number of moves and value of the treasure, the quality of a state can be assessed by two rewards, $\rho_p$ and $\rho_v$, for each objective respectively. $\rho_p$ is adjusted to be maximized using the maximum number of moves in the game, while $\rho_v$ is simply the value of the cell that holds the treasure. Therefore, the reward vector to maximize is defined as $\overline{r} = \{ \rho_p, \rho_v\}$. Equation~\ref{eq:dst} summarizes these rewards:

@@

This needs to be accomplished while optimizing three different objectives:  1) \textbf{Time}: the player must collect all waypoints scattered around the maze in as few time steps as possible; 2) \textbf{Fuel}: the fuel consumption by the end of the game must be minimized; and 3) \textbf{Damage}: the ship should end the game with as little damage as possible.

@@

Regarding the third objective, the ship can suffer damage in two different ways: by colliding with obstacles and driving through lava. In the former case, the ship can collide with normal obstacles (that subtract $10$ units of damage) and especially damaging obstacles ($30$ units). In the latter, lava lakes are abundant in MO-PTSP levels and, in contrast with normal surfaces, they deal one unit of damage for each time step the ship spends over this type of surface. All these subtractions are deducted from an initial counter of $5000$ points of damage.

@@

All the algorithms tested in the MO-PTSP in this research employ macro-actions, a concept that can be used for coarsening the action space by applying the action chosen by the control algorithm in several consecutive cycles, instead of just in the next one.

@@

$\lambda_t$ is the fuel consumed so far, and $\lambda_0$ is the initial fuel at the start of the game. $\alpha$ is a value that balances between the fuel component and the time objective from Equation~\ref{eq:dist}. Note that, as waypoints need to keep being visited, it is necessary to include a distance measure for this reward. Otherwise, an approach that prioritizes this objective would not minimize distance to waypoints at all (the ship could just stand still: no fuel consumed is optimal), and therefore would not complete the game. The value of $\alpha$ has been determined empirically, in order to provide a good balance between these two components, and is set to $0.66$.

Finally, Equation~\ref{eq:dam} gives the method used to calculate the damage objective, $\rho_d$:

@@

$g_t$ is the damage suffered so far in the game, and $g_M$ is the maximum damage the ship can take. In this case, three different variables are used to regulate the behaviour of this objective: $\gamma$, $\beta_{1}$ and $\beta_{2}$. Both $\beta_{1}$ and $\beta_{2}$ have the same role as $\alpha$ in Equation~\ref{eq:fuel}: they balance between the time objective and the damage measure. The difference is that $\beta_{1}$ is used in high speeds, while $\beta_{2}$ is employed with low velocities. This is distinguished by the parameter $\gamma$, that can be seen as a threshold value for the ship's speed ($sp$). This differentiation is made in order to avoid low speeds in lava lakes, as this significantly increases the damage suffered. The values for these variables have also been determined empirically, and they are set to $\gamma = 0.8$, $\beta_{1} = 0.75$ and $\beta_{2} = 0.25$.

The final reward vector to be maximized is therefore $\overline{r} = \{ \rho_t, \rho_f, \rho_d \}$. It is important to highlight again that these three rewards are the same for all the algorithms tested in the experiments.

@@

The experiments performed in this research compare three different algorithms in the two benchmarks presented in Section~\ref{sec:bench}: a single objective MCTS (referred to here simply as \textit{MCTS}), the Multi-Objective MCTS (\textit{MO-MCTS}) and a rolling horizon version of the NSGA-II algorithm described in Section~\ref{sec:moo} (\textit{NSGA-II}). This NSGA-II version evolves a population where the individuals are sequences of actions (macro-actions in the MO-PTSP case), obtaining the fitness from the state of the game after applying the sequence. The population sizes, determined empirically, were set to $20$ and $50$ individuals for DST and MO-PTSP respectively. The value of $C$ in Equation~\ref{eq:ucb1} and~\ref{eq:moucb} is set to $\sqrt{2}$.

All the algorithms have a limited number of evaluations before providing an action to perform. In order for these games to be real-time, the time budget allowed is close to $40$ milliseconds. With the objective of avoiding congestion peaks at the machine where the experiments are run, the average number of evaluations possible in $40$ ms is calculated and employed in the tests. This leads to $4500$ evaluations in the DST, and $500$ evaluations for MO-PTSP, using the same server where the PTSP and MO-PTSP competitions were run\footnote{Intel Core i5, 230GHz, 4GB of RAM.}.

@@

The results show clearly how approximating the optimal Pareto front allows for finding all possible solutions. Both the NSGA-II and MO-MCTS approaches are able to converge to any solution in the front given the appropriate weight vector. It is important to highlight that these two algorithms employed the Euclidean distance action selection (see Section~\ref{sec:momcts}). Other experiments, not included in this paper, showed that weighted sum action selection provides similar results to MCTS (that is, convergence to the edges of the front). The crucial distinction to make here is that both algorithms, NSGA-II and MO-MCTS, allow for better action selection mechanisms, which are able to overcome this problem by approximating a global Pareto front.

@@

The same three algorithms have been tested in the MO-PTSP domain. The experiments are performed in the $10$ maps available within the framework~\footnote{Available at \url{www.ptsp-game.net}}. In this case, the optimal Pareto front is not known in advance, and normally differs from one map to another. Hence, the mechanisms to compare the performance of the algorithms tested need to be different than the one used for DST.

The idea is as follows. First of all, $4$ different weight vectors are tested: $W_1 = (0.33,0.33,0.33)$, $W_2 = (0.1,0.3,0.6)$, $W_3 = (0.1,0.6,0.3)$ and $W_4 = (0.6,0.1,0.3)$, where each $w_i$ corresponds to the weight for an objective ($w_t$ for time, $w_f$ for fuel and $w_d$ for damage, in this order). $W_1$ treats all objectives as having the same weight, while the other three give more relevance to different objectives. These vectors then provide a wide spectrum for weights in this benchmark. In this case, MO-MCTS uses these weights to select an action based on a weighted sum, which in preliminary experiments has shown better performance than the Euclidean distance mechanism.

@@

Some exceptions can be explained by analysing the specific maps: in map 1, a fuel canister is always collected near the end, restoring the fuel level to its maximum. This leaves too few cycles to make a difference in the controller (note also that this is the map with the smallest fuel consumption). Also, in map 10, there is no difference in the damage objective, however map 10 is a map with no obstacles to damage the ship (thus only lava lakes deal damage). This also results in this map being the one with the lowest overall damage. It can also be seen that, in those maps where time has priority, the results for this objective are not as dominant as the other two. This can be explained by the fact that the time heuristic is actually part of the fuel and damage heuristics (Equation~\ref{eq:dist} in Eqs.~\ref{eq:fuel} and~\ref{eq:dam}).

@@

Extending this comparison to all maps, each pair of algorithms ends with a triplet $(D, \O, d)$, where $D$ is the number of maps where the first algorithm dominates the second, $\O$ is the amount of maps where no dominance can be established, and $d$ states the number of maps where the first is dominated by the second. Table~\ref{tab:resMOPTSP} summarizes these results for all the algorithms tested.

\begin{table*}[!t]
\begin{center}
\begin{tabular}{|c|c|c|c|c|c|}
\hline
  & \textbf{$W: (w_t, w_f, w_d)$} & \textbf{MO-MCTS} \bm{$(D, \O, d)$} &  \textbf{MCTS} \bm{$(D, \O, d)$} & \textbf{NSGA-II} \bm{$(D, \O, d)$} & \textbf{PurofMovio} \bm{$(D, \O, d)$} \\ 
\hline
 \multirow{3}{*}{\textbf{MO-MCTS}} & $W_1: (0.33,0.33,0.33)$ & \multirow{3}{*}{$-$} & $(8,2,0)$ & $(8,2,0)$ & $(0,5,5)$\\
 & $W_2: (0.1,0.3,0.6)$ & & $(10,0,0)$ & $(4,6,0)$ & $(0,6,4)$\\
 & $W_3: (0.1,0.6,0.3)$ & & $(8,2,0)$ & $(7,3,0)$ & $(0,5,5)$\\
 & $W_4: (0.6,0.1,0.3)$ & & $(10,0,0)$ & $(3,7,0)$ & $(0,3,7)$\\
\hline
 \multirow{3}{*}{\textbf{MCTS}} & $W_1: (0.33,0.33,0.33)$& $(0,8,2)$ & \multirow{3}{*}{$-$} & $(0,2,8)$ & $(0,2,8)$\\
 & $W_2: (0.1,0.3,0.6)$ & $(0,0,10)$ & & $(4,2,4)$ & $(0,3,7)$\\
 & $W_3: (0.1,0.6,0.3)$ & $(0,2,8)$ & & $(0,1,9)$ & $(0,6,4)$\\
 & $W_4: (0.6,0.1,0.3)$ & $(0,0,10)$ &  & $(3,3,4)$ & $(0,1,9)$\\
\hline
 \multirow{3}{*}{\textbf{NSGA-II}} & $W_1: (0.33,0.33,0.33)$& $(0,2,8)$ & $(8,2,0)$ & \multirow{3}{*}{$-$} & $(0,4,6)$\\
 & $W_2: (0.1,0.3,0.6)$ & $(0,6,4)$ & $(4,2,4)$ & & $(0,4,6)$\\
 & $W_3: (0.1,0.6,0.3)$ & $(0,3,7)$ & $(9,1,0)$ & & $(0,5,5)$\\
 & $W_4: (0.6,0.1,0.3)$ & $(0,7,3)$ & $(4,3,3)$ &  & $(0,4,6)$\\
\hline
 \multirow{3}{*}{\textbf{PurofMovio}} & $W_1: (0.33,0.33,0.33)$& $(5,5,0)$ & $(8,2,0)$ & $(6,4,0)$ & \multirow{3}{*}{$-$}\\
 & $W_2: (0.1,0.6,0.3)$ & $(4,6,0)$ & $(7,3,0)$ & $(6,4,0)$ & \\
 & $W_3: (0.1,0.3,0.6)$ & $(5,5,0)$ & $(6,4,0)$ & $(5,5,0)$ & \\
 & $W_4: (0.6,0.1,0.3)$ & $(7,3,0)$ & $(9,1,0)$ & $(6,4,0)$ & \\
\hline
\end{tabular}
\caption{Results in MO-PTSP: Each cell indicates the triplet $(D, \O, d)$, where $D$ is the number of maps where the row algorithm dominates the column one, $\O$ is the amount of maps where no dominance can be established, and $d$ states the number of maps where the row algorithm is dominated by the column one. All the algorithms followed the same route (order of waypoints and fuel canisters) in every map tested.}
\label{tab:resMOPTSP}
\end{center}
\end{table*}

One of the first things to notice is that MO-MCTS dominates MCTS and NSGA-II in most of the maps, and it is never dominated in any. In particular, the dominance of MO-MCTS over MCTS is outstanding, even dominating in all $10$ maps for two of the weight vectors. MO-MCTS also dominates NSGA-II in more maps than in those where there is no dominance, and  is never dominated by NSGA-II in any map.

@@

There is another aspect that can be further improved in the MO-PTSP benchmark, and is also applicable to other domains. It is naive to think that a unique weight vector will be the ideal one for the whole game. Specifically in the MO-PTSP, there are regions of the map where there are more obstacles or lava lakes, hence the ship is most likely to suffer higher damage there. Also, the route followed during the game affects the relative ideal speed between waypoints, or perhaps a fuel canister may be picked up, which will affect how the fuel objective will be managed. In general, many real-time games go through different phases, with different objectives and priorities.
